---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  out.width = "100%"
)
```

# libtiledbvcf and TileDB-VCF CLI For R

Minimal R bindings and CLI wrapper for [TileDB-VCF](https://github.com/TileDB-Inc/TileDB-VCF).

We build the CLI and the library, provide a CLI wrapper. CLI wrapped using processx is used for creating, ingesting, exporting, and querying TileDB-VCF datasets and library will be used for future R bindings. We may add some additional functionality using vcfppR (processx callbacks for example). This is a work in progress.


## Installation

This package build from source htslib among other dependencies of libtiledbvcf per the official cmake build instructions. This setup requires autoreconf and an up to date cmake sadly, so cmake, autoreconf, neeeds to be installed on your system. In future we may use [`biocmake`](https://github.com/LTLA/biocmake/) package to make sure cmake is downloaded if needed. We may find the way to reliably use external htlib to link against but this look brittle. So for macOS you can use `brew install cmake autoconf automake libtool` to get started (as the github action of this package does it) . For linux you can use your package manager to install these dependencies (they are usally installed). For windows we recommend using WSL2 with Ubuntu since we have no plans to configure the package for windows.


## Setup and Version Checking


```{r, eval = TRUE}
library(Rlibtiledbvcf)
library(vcfppR)
tiledb_vcf_cli_version() |> cat(sep = "\n")
tiledb_vcf_version() |> cat(sep = "\n")
```

## Creating a TileDB-VCF Dataset

```{r, eval = TRUE}
# Create a new dataset
uri <- file.path(tempdir(), "my_dataset")
tiledb_vcf_create(uri, print_command = TRUE)
```

##  Ingesting Multiple VCFs

```{r, eval = TRUE}
sample_names <- c("HG00096", "HG00097", "HG00099", "HG00100")
sample_vcfs <- vapply(
    sample_names,
    function(s) {
        system.file("extdata", paste0(s, ".vcf.gz"), package = "RTileDBvcf")
    },
    character(1)
)

# Write samples file for ingestion (one VCF per line)
samples_file <- tempfile(fileext = ".txt")
writeLines(sample_vcfs, samples_file)

# Ingest using TileDB-VCF CLI
tiledb_vcf_store(
    uri,
    samples_file = samples_file,
    echo = TRUE,
    print_command = TRUE
)
```

## Listing Samples in the Dataset

```{r, eval = TRUE}
tiledb_vcf_list(uri)
```

## Exporting and Validating VCFs (Round-Trip)

```{r, eval = TRUE}
# Export and compare each sample individually
all_identical <- TRUE
for (i in seq_along(sample_names)) {
    export_file <- tempfile(fileext = ".vcf")
    tiledb_vcf_export(
        uri = uri,
        output_path = export_file,
        sample_names = sample_names[i], # Export one sample at a time
        merge = TRUE,
        echo = TRUE
    )
    stopifnot(file.exists(export_file))
    orig_tab <- vcftable(sample_vcfs[i])
    exp_tab <- vcftable(export_file)
    # Filter exported table for current sample if needed
    if ("sample" %in% names(exp_tab)) {
        exp_tab <- exp_tab[exp_tab$sample == sample_names[i], ]
    }
    # Compare relevant columns
    identical_chr <- identical(orig_tab$chr, exp_tab$chr)
    identical_pos <- identical(orig_tab$pos, exp_tab$pos)
    identical_gt <- identical(orig_tab$gt, exp_tab$gt)
    if (!(identical_chr && identical_pos && identical_gt)) {
        all_identical <- FALSE
        cat("Comparison failed for sample:", sample_names[i], "\n")
    }
}
if (all_identical) {
    cat("All samples are identical in chrom, pos, and GT columns.\n")
}
```


## References

- [TileDB-VCF GitHub Repository](https://github.com/TileDB-Inc/TileDB-VCF)